Leadership across cultures and contexts: an exploratory analysis

### 2020-05-09 leadership_across_cultures_contexts.Rmd
Revising discussion. Spell check document.

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*.

@@

The distribution of text records by group context and subsistence strategy suggests leaders emerge within similar social contexts across populations, but at variable frequencies. See Figure \@ref(fig:grouptype). Leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013]. Supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support a relationship between residential mobility and subsistence intensification with sociopolitical organization among non-WEIRD populations. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

One possible interpretation of the negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain important leadership positions within the groups and represent the group to governmental forces. @aa01-062 [p. 173] describes that in these local 'clubs',

>"the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among nonhuman animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

Leadership is expected to be associated with particular suites of characteristics and functions, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @van_vugt_dual_2019], and as discussed, systematic cross-cultural evidence here supports many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and inspired much subsequent theoretical and empirical work. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as two distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, in order to organize group members, impliment strategic actions, and provide prosocial services to the group, all while conforming to social and cultural pressures.

>>>>

Words that were strong predictors of leaders with coercive authority included, *power* and *authority* (unsurprisingly), as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. See Figure \@ref(fig:elasticcoercive)A. Using all other quality and function variables as predictors identified *Punishment* as the strongest predictor of evidence for *Coercive authority*. The qualities of *Aggressiveness* and *Feared* were also positive predictors, as were the functions, *Political appointments*, *Control economics*, and *Control immigration*. Furthermore, evidence for leaders lacking coercive authority (our *No coercive authority* variable) was documented in over 35% of cultures (Figure \@ref(fig:varsupport)) and demonstrated variability by susbsitence strategy with hunter-gatherers providing relaitvely higher levels of evidence for leaders lacking coercive authority. See Figure S\@ref(fig:heatmapdrop1).

@@

The distribution of text records by group context and subsistence strategy suggests that although leaders emerge within similar social contexts across populations, it occurs at variable frequencies. See Figure \@ref(fig:grouptype). Leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories. These patterns support perspectives suggesting residential mobility and increased costs of dispersal, subsistence intensification, and demographic transitions are associated with increasing complexity of sociopolitical organization among non-WEIRD populations [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section). 
<!-- Thus, these results provide a fine-grained empirical view of putative cultural transitions in leadership systems. -->

Pastoralists, despite potential for residential mobility, do rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 [p. 58] notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

One possible interpretation of the negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level -- more similar to leadership systems of WEIRD populations -- leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintains important leadership positions within the group and represents the group to governmental bodies @aa01-062 [p. 173] describes that in these local 'clubs',

>the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role.

Several leadership dimensions that are theoretically important were relatively rare in the ethnographic record. The leadership literature among nonhuman animals, for example, commonly focuses on group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among small-scale human societies and nonhuman animal societies, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 [p. 411] describes that, "if there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

Leadership and social influence expected to be associated with particular suites of characteristics and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990]. Our systematic cross-cultural evidence supports many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social heirarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_j.k_dominance_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, impliment strategic actions, and provide prosocial services to the group, all while conforming to cultural pressures. Compartamentalizing this range of diversity into two distinct social stratiges, although theoretically and empirically convenient, likely misses much of the ethnographic reality of human leadership dynamics.
